Sequential rank agreement methods for comparison of ranked lists

### 2016-05-23 latex/superranker-manuscript.tex
proof read proof

@@

\section{Asymptotics}\label{sec:chgpoint}
In many applications it is of interest to estimate a list depths which
satisfies a change point criterion. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As Hall and Schimek \mcomment{REF}, we work with an infinite set of lists
and study the asymptotic behaviour for $L\to\infty$.
However, we do not allow the length of the lists to
change with $L$ as in most applications the length of the lists is fixed.

We start by showing that $\widehat{\textrm{sra}}_L$ is a consistent estimator of $\textrm{sra}$ for $L \rightarrow \infty$.
The proof is given in Theorem \ref{thm:consistency} below and requires the following assumptions.
\begin{assumption}
We assume that:
\begin{enumerate}[label=\Alph*)]
\item{The $L$ lists are independent and identically distributed samples from $Q$.}
\item{$\E_Q[R(X_p)^2] < \infty$ uniformly in $p$.}
\item{$|S(d)| > 0$ uniformly in $d$}.
\end{enumerate}
\end{assumption}
We note that assumption $C$ implies that for every list depth $d$ there must be a non-zero probability of ... under the probability measure $Q$.

\begin{theorem}
Under the assumptions it follows that $\left\|\widehat{\textrm{sra}}_L - \textrm{sra}\right\|_\infty = o_P(1)$.
\label{thm:consistency}
\end{theorem}

@@

\newpage
\section{Proof of Theorem \ref{thm:consistency}}
We start by defining the following cadlag function
\begin{align}
  N(p; d) = \sum_{s=1}^p 1\left(Q(R(X_s) \leq d) > 0\right), \quad p = 1,\ldots, P
\end{align}
which jumps with an increment of one at time $p$ if the probability that the rank of $X_p$ is less than or equal to $d$ is bounded away from zero under the probability measure $Q$.
Similarly, we have the empirical counterpart
\begin{align}
  \widehat{N}_L(p; d) &= \sum_{s=1}^p 1\left( \frac{1}{L} \sum_{l=1}^L 1(R_l(X_s) \leq d) > 0\right)\\
        &= \sum_{s=1}^p 1\left(\widehat{Q}_L(R(X_s) \leq d) > 0\right)\nn
\end{align}
The joint law of $\{\widehat{N}_L(p;d)\}_{1\leq p \leq P}$ is completely determined by its finite set of jump times, and because
$1(R_l(X_s) \leq d)$ are independent and identically Bernoulli distributed over $l$ with expectation $Q(R(X_s) \leq d)$ 
it follows from the law of large numbers that $\widehat{Q}_L(R(X_p) \leq d) \overset{P}{\longrightarrow} Q(R(X_p) \leq d)$ for every $p$ and $d$ 
for $L \rightarrow \infty$, and therefore
\begin{align}
  \sup_{d \in 1,\ldots,P}\sup_{p \in 1,\ldots, P} \left|\widehat{N}_L(p; d) - N(p; d)\right| = o_P(1)\label{eq:countingConv}
\end{align}

The sequential rank agreement given in equation (\ref{def:sra}) may therefore be rewritten as the following integral with respect to the counting measure induced by $N$
\begin{align}
  \textrm{sra}(d) = \int_{1}^P \frac{A(X_p)}{N(P; d)}dN(p; d)
\end{align}
where we note that the cardinality of the set $S(d)$ in equation (\ref{eq:sumset}) is equal to the counting process evaluated at time $P$.
The empirical sequential rank agreement is similarly given by
\begin{align}
  \widehat{\textrm{sra}}_L(d) &= \int_1^P \frac{\widehat{A}_L(X_p)}{\widehat{N}_L(P; d)}d\widehat{N}_L(p; d) 
\end{align}
and by elementary algebra it follows that
\begin{align}
  \sup_{d \in 1,\ldots,P}\left|\widehat{\textrm{sra}}_L(d) - \textrm{sra}(d)\right| \leq \sup_{d \in 1,\ldots,P}\left|U_L(d)\right| - \sup_{d \in 1,\ldots,P}\left|V_L(d)\right|
\end{align}
where
\begin{align}
  U_L(d) &= \int_1^P\frac{A(X_p)}{N(P;d)} d\left(\widehat{N}_L(p; d) - N(p; d)\right)\\
  V_L(d) &=  \int_1^P \left(\frac{\widehat{A}_L(X_p)}{\widehat{N}_L(P; d)} - \frac{A(X_p)}{N(P;d)}\right)d\widehat{N}_L(p; d)
\end{align}
The conclusion of the proof therefore follows if each of these two terms are uniformly of order $o_P(1)$ in $d$.

Starting with the first term, we have that
\begin{align}
  \left|U_L(d)\right| \leq \left(\sup_{p \in 1,\ldots,P}\left|\frac{A(X_p)}{N(P;d)}\right|\right)\left(\widehat{N}_L(p; d) - N(p; d)\right)\biggr\rvert_1^P
\end{align}
where the second factor is $o_P(1)$ uniformly in $d$ by equation (\ref{eq:countingConv}). The first factor is $O(1)$ since $N(P;d)$ uniformly bounded away from zero in d by assumption C and $A(X_p)$ is uniformly bounded in $p$ by assumption B. It thus follows that $\left|U_L(d)\right| = o_P(1)$ uniformly in $d$.

Continuing with $V_L(d)$ we may similarly bound as
\begin{align}
 V_L(d) &\leq \sup_{p \in 1,\ldots,P}\left|\frac{\widehat{A}_L(X_p)}{\widehat{N}_L(P; d)} - \frac{A(X_p)}{N(P;d)}\right| \int_1^P \left|d\widehat{N}_L(p; d)\right|\label{eq:secondTerm}
\end{align}
It follows directly from the law of large numbers that $\widehat{A_L}(X_p) = A(X_p) + o_P(1)$ uniformly in $p$ since $R_i(X_p)$ in equation (\ref{eq:empDistance}) for $i=1,\ldots,L$ are independent and identically distributed random variables from $Q$ with finite second moment by assumption B). Again by equation (\ref{eq:countingConv}) it follows that $\widehat{N}_L(P; d) = N(P;d) + o_P(1)$ uniformly in $d$ and therefore by the continuous mapping theorem and assumption C) the first factor in equation (\ref{eq:secondTerm}) is of order $o_P(1)$ uniformly in $p$ and $d$. The result then follows by noting that the second factor is bounded by $P$ uniformly in $d$.

This completes the proof.

>>>>

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Fri Apr  8 10:22:58 2016
\begin{table}[tb]
\centering
\caption{Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method} 
\label{tab1}
\begin{tabular}{rrrrr}
  \hline
Ranking & T & LogReg & ElasticNet & MIC \\ 
  \hline
1 & 2124 & 2124 & 829 & 378 \\ 
  2 & 896 & 896 & 2124 & 829 \\ 
  3 & 2600 & 829 & 2198 & 896 \\ 
  4 & 766 & 394 & 1665 & 1037 \\ 
  5 & 829 & 766 & 1920 & 2124 \\ 
  6 & 2851 & 2670 & 1042 & 808 \\ 
  7 & 703 & 2939 & 808 & 108 \\ 
  8 & 2386 & 2386 & 849 & 515 \\ 
  9 & 2645 & 1834 & 937 & 2670 \\ 
  10 & 2002 & 378 & 1995 & 2600 \\ 
   \hline
\end{tabular}
\end{table}

@@

\section{Asymptotics}\label{sec:chgpoint}
In many applications it is of interest to estimate a list depths which
satisfies a change point criterion. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As Hall and Schimek \mcomment{REF}, we work with an infinite set
of lists and study the asymptotic behaviour for $L\to\infty$.
However, we do not allow the length of the lists to change with $L$
because in most applications the lengths of the lists are fixed.

We start by showing that $\widehat{\textrm{sra}}_L$ is a consistent
estimator of $\textrm{sra}$ for $L \rightarrow \infty$. The result is
stated in Theorem \ref{thm:consistency}.

% below and requires the
% following assumptions.
% \begin{assumption}
% We assume that:
% \begin{enumerate}[label=\Alph*)]
% \item{The $L$ lists are independent and identically distributed
    % samples from a probability distribution $Q$ on the set of lists.}
% \item{$\E_Q[R(X_p)^2] < \infty$ uniformly in $p$.}
% \item{$|S(d)| > 0$ uniformly in $d$}.
% \end{enumerate}
% \end{assumption}
% We note that assumption $C$ implies that for every list depth $d$
% there must be a non-zero probability of ... under the probability
% measure $Q$.

\begin{theorem}
  Assume that $\{R_l(X)\}_{l=1}^L$ are independent draws from the a
  probability distribution $Q$ on the set of lists.  Then,
  $\left\|\widehat{\textrm{sra}}_L - \textrm{sra}\right\|_\infty =
  o_P(1)$.
\label{thm:consistency}
\end{theorem}

@@

\newpage
\section{Proof of Theorem \ref{thm:consistency}}

We start by defining the following counting process
\begin{align}
  N(p; d) = \sum_{s=1}^p 1\left(Q(R(X_s) \leq d) > 0\right), \quad p = 1,\ldots, P
\end{align}
which runs through the list elements in an arbitrary order and counts
how many list elements have a strictly positive probability to have
rank less than or equal to $d$ under the probability measure $Q$.
Note that the cardinality of the set $S(d)$ in equation
(\ref{eq:sumset}) is equal to the counting process evaluated at the last element:
\begin{equation*}
  |S(d)| = N(P;d).
\end{equation*}

The empirical counterpart of the counting process is given by
\begin{align}
  \widehat{N}_L(p; d) &= \sum_{s=1}^p 1\left( \frac{1}{L} \sum_{l=1}^L 1(R_l(X_s) \leq d) > 0\right)\\
        &= \sum_{s=1}^p 1\left(\widehat{Q}_L(R(X_s) \leq d) > 0\right)\nn.
\end{align}
The joint law of the sequence
$\{\widehat{N}_L(p;d)\}_{1\leq p \leq P}$ is completely determined by
the finite set of the jump times of $N(p;d)$. Since
$\{1(R_l(X_s) \leq d); l=1,\dots, L\}$ consists of independent and
identically Bernoulli distributed variables with expectation
$Q(R(X_s) \leq d)$ it follows from the law of large numbers that
$\widehat{Q}_L(R(X_p) \leq d) \overset{P}{\longrightarrow} Q(R(X_p)
\leq d)$ for every $p$ and $d$ as $L \rightarrow \infty$. 
Therefore, we have
\begin{align}
  \sup_{d \in 1,\ldots,P}\sup_{p \in 1,\ldots, P} \left|\widehat{N}_L(p; d) - N(p; d)\right| = o_P(1).\label{eq:countingConv}
\end{align}

The sequential rank agreement given in equation
(\ref{def:sra}) may therefore be rewritten as the following integral
with respect to $N$
\begin{align}
  \textrm{sra}(d) = \int_{1}^P \frac{A(X_p)}{N(P; d)}\mathrm dN(p; d).
\end{align}
The empirical sequential rank agreement is similarly given by
\begin{align}
  \widehat{\textrm{sra}}_L(d) &= \int_1^P \frac{\widehat{A}_L(X_p)}{\widehat{N}_L(P; d)}\mathrm d \widehat{N}_L(p; d) 
\end{align}
and it follows that
\begin{align}
  \sup_{d \in 1,\ldots,P}\left|\widehat{\textrm{sra}}_L(d) - \textrm{sra}(d)\right| \leq \sup_{d \in 1,\ldots,P}\left|U_L(d)\right| - \sup_{d \in 1,\ldots,P}\left|V_L(d)\right|
\end{align}
where
\begin{align}
  U_L(d) &= \int_1^P\frac{A(X_p)}{N(P;d)} \mathrm d\left(\widehat{N}_L(p; d) - N(p; d)\right)\\
  V_L(d) &=  \int_1^P \left(\frac{\widehat{A}_L(X_p)}{\widehat{N}_L(P; d)} - \frac{A(X_p)}{N(P;d)}\right)\mathrm d\widehat{N}_L(p; d)
\end{align}
The conclusion of the proof therefore follows if each of these two
terms are uniformly of order $o_P(1)$ in $d$.  For the first term we
have that
\begin{align}
  \left|U_L(d)\right| \leq \left(\sup_{p \in 1,\ldots,P}\left|\frac{A(X_p)}{N(P;d)}\right|\right)\left(\widehat{N}_L(p; d) - N(p; d)\right)\biggr\rvert_1^P
\end{align}
where the second factor is $o_P(1)$ uniformly in $d$ by equation
(\ref{eq:countingConv}). The first factor is $O(1)$ since $N(P;d)>0$
uniformly in $d$ and because $A(X_p)\le 2L^2$ u uniformly in $p$. . It
thus follows that $\left|U_L(d)\right| = o_P(1)$ uniformly in $d$.

Similarly we derive an upper bound for $V_L(d)$ by
\begin{align}
 V_L(d) &\leq \sup_{p \in 1,\ldots,P}\left|\frac{\widehat{A}_L(X_p)}{\widehat{N}_L(P; d)} - \frac{A(X_p)}{N(P;d)}\right| \int_1^P \left|\mathrm d\widehat{N}_L(p; d)\right|\label{eq:secondTerm}
\end{align}
It follows directly from the law of large numbers that
$\widehat{A_L}(X_p) = A(X_p) + o_P(1)$ uniformly in $p$ since
$R_i(X_p)$ in equation (\ref{eq:empDistance}) for $i=1,\ldots,L$ are
independent and $Q$-identically distributed random variables with
finite second moment since $A(X_p)\le 2L^2$). Again by equation
(\ref{eq:countingConv}) it follows that
$\widehat{N}_L(P; d) = N(P;d) + o_P(1)$ uniformly in $d$ and therefore
by the continuous mapping theorem and since $N(P;d)>0$) the first factor
in equation (\ref{eq:secondTerm}) is of order $o_P(1)$ uniformly in
$p$ and $d$. The result then follows by noting that the second factor
is bounded by $P$ uniformly in $d$.
% This completes the proof.
